---
title: "Using fz with Modelica Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using fz with Modelica Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when fz Python package is properly configured
)
```

```{r setup}
library(fz)
```

## Introduction

This vignette demonstrates how to use the `fz` R package to run design of
experiments, optimization, and uncertainty quantification with Modelica models.

## Installation

First, ensure you have the fz Python package installed:

```{r install}
# Install fz Python package
fz_install()

# Verify installation
fz_available()
```

## Basic Workflow

The typical workflow with fz involves:

1. **Initialize** a project with a Modelica model (`fzi`)
2. **Configure** input variables and outputs (`fzc`)
3. **Execute** design of experiments or optimization (`fzd`, `fzo`, or `fz`)
4. **Analyze** results

## Example 1: Bouncing Ball Simulation

This example demonstrates a simple design of experiments with a bouncing ball
Modelica model.

### Model Description

The bouncing ball model simulates a ball dropping from an initial height with
gravity and bouncing with a coefficient of restitution.

**Input variables:**
- `h0`: Initial height (m)
- `v0`: Initial velocity (m/s)
- `e`: Coefficient of restitution (0-1)

**Output variables:**
- `h_max`: Maximum bounce height (m)
- `t_ground`: Time to first ground contact (s)

### Running the Design of Experiments

```{r bouncing_ball}
# Initialize project with Modelica model
fzi(
  model = "modelica",
  model_path = "path/to/BouncingBall.mo"
)

# Configure input variables with ranges
fzc(
  input = list(
    h0 = c(1.0, 10.0),    # height from 1m to 10m
    v0 = c(-2.0, 2.0),    # velocity from -2 to 2 m/s
    e = c(0.5, 0.9)       # restitution from 0.5 to 0.9
  ),
  output = c("h_max", "t_ground")
)

# Run Latin Hypercube sampling with 50 samples
results <- fzd(
  design = "LatinHypercube",
  n = 50
)

# Analyze results
summary(results)
plot(results$h0, results$h_max,
     xlab = "Initial Height (m)",
     ylab = "Maximum Bounce Height (m)")
```

## Example 2: Spring-Mass-Damper Optimization

This example shows how to optimize parameters of a spring-mass-damper system.

### Model Description

A spring-mass-damper system responding to an initial force.

**Input variables:**
- `m`: Mass (kg)
- `k`: Spring stiffness (N/m)
- `c`: Damping coefficient (N·s/m)

**Output variables:**
- `settling_time`: Time to settle (s)
- `overshoot`: Maximum overshoot (%)

### Running the Optimization

```{r spring_mass}
# Initialize
fzi(
  model = "modelica",
  model_path = "path/to/SpringMassDamper.mo"
)

# Configure with fixed force
fzc(
  input = list(
    m = c(0.5, 5.0),
    k = c(100, 10000),
    c = c(1, 100)
  ),
  fixed = list(F0 = 100),  # Fixed initial force
  output = c("settling_time", "overshoot")
)

# Optimize to minimize settling time while constraining overshoot
results <- fzo(
  objective = "minimize",
  objective_var = "settling_time",
  constraints = list(overshoot = c(-Inf, 10)),  # Max 10% overshoot
  algorithm = "GradientDescent"
)

# View optimal parameters
print(results$optimal_input)
print(results$optimal_output)
```

## Example 3: Heat Exchanger Parameter Study

This example demonstrates a full factorial design for a heat exchanger.

### Model Description

A counter-flow heat exchanger with hot and cold fluid streams.

**Input variables:**
- `mdot_hot`: Hot fluid flow rate (kg/s)
- `mdot_cold`: Cold fluid flow rate (kg/s)

**Output variables:**
- `effectiveness`: Heat exchanger effectiveness (0-1)
- `Q_total`: Total heat transfer (W)

### Running the Parameter Study

```{r heat_exchanger}
# Initialize
fzi(
  model = "modelica",
  model_path = "path/to/HeatExchanger.mo"
)

# Configure variables
fzc(
  input = list(
    mdot_hot = c(0.5, 1.0, 1.5),   # 3 levels
    mdot_cold = c(0.5, 1.0, 1.5)   # 3 levels
  ),
  fixed = list(
    T_hot_in = 80,
    T_cold_in = 20
  ),
  output = c("effectiveness", "Q_total")
)

# Full factorial design: 3 x 3 = 9 runs
results <- fzd(
  design = "FullFactorial"
)

# Create response surface
library(ggplot2)
ggplot(results, aes(x = mdot_hot, y = mdot_cold, fill = effectiveness)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Heat Exchanger Effectiveness",
    x = "Hot Flow Rate (kg/s)",
    y = "Cold Flow Rate (kg/s)"
  )
```

## Example 4: Uncertainty Quantification

This example shows how to perform uncertainty quantification with uncertain
input parameters.

```{r uncertainty}
# Initialize model
fzi(
  model = "modelica",
  model_path = "path/to/System.mo"
)

# Configure with probabilistic inputs
fzc(
  input_dist = list(
    friction_coef = list(
      type = "normal",
      mean = 0.3,
      sd = 0.05
    ),
    ambient_temp = list(
      type = "uniform",
      min = 15,
      max = 35
    )
  ),
  output = "performance"
)

# Run Monte Carlo simulation
results <- fz(
  n_samples = 1000,
  method = "MonteCarlo"
)

# Analyze uncertainty in output
hist(results$performance,
     main = "Performance Distribution",
     xlab = "Performance",
     breaks = 30)

# Summary statistics
cat("Mean:", mean(results$performance), "\n")
cat("Std Dev:", sd(results$performance), "\n")
cat("95% CI:", quantile(results$performance, c(0.025, 0.975)), "\n")
```

## Advanced Usage

### Custom Design Algorithms

```{r custom_design}
# Different DoE algorithms can be used with fzd
# Latin Hypercube Sampling - good space-filling properties
results_lhs <- fzd(design = "LatinHypercube", n = 100)

# Sobol sequences - quasi-random low-discrepancy
results_sobol <- fzd(design = "Sobol", n = 100)

# Random sampling
results_random <- fzd(design = "Random", n = 100)

# Full factorial (uses all combinations)
results_factorial <- fzd(design = "FullFactorial")
```

### Parallel Execution

```{r parallel}
# Enable parallel execution for faster computation
fzc(
  parallel = TRUE,
  n_cores = 4
)
```

## Best Practices

1. **Start small**: Test with a few samples before running large DoE
2. **Validate models**: Ensure Modelica models compile and run independently
3. **Check ranges**: Verify input ranges are physically meaningful
4. **Monitor convergence**: For optimization, check convergence criteria
5. **Save results**: Store results for reproducibility

```{r save_results}
# Save results to file
saveRDS(results, "fz_results.rds")

# Export to CSV for analysis in other tools
write.csv(results, "fz_results.csv", row.names = FALSE)
```

## Troubleshooting

### Common Issues

**Python package not found:**
```{r troubleshoot1}
# Reinstall fz Python package
fz_install()
```

**Modelica compilation errors:**
- Check that the Modelica model path is correct
- Ensure the model compiles in your Modelica environment
- Verify all dependencies are available

**Slow execution:**
- Reduce number of samples
- Enable parallel execution
- Use faster optimization algorithms for initial exploration

## Further Reading

- Funz documentation: [https://funz.github.io](https://funz.github.io)
- Modelica language: [https://modelica.org](https://modelica.org)
- Design of Experiments: See references on DoE methods

## Session Info

```{r session_info}
sessionInfo()
```
